---
title: "Methodology Guide"
author: "Avishek Bhandari"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Methodology Guide}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6.5,
  fig.height = 4
)
```

# Overview

`contagionchannels` operationalises a two-stage research design that is
deliberately ecumenical about identification. Stage 1 *detects*
directional information flow between markets using Wavelet-Quantile
Transfer Entropy (WQTE). Stage 2 *attributes* the detected flow to five
economically meaningful channels using a battery of estimators that lean
on different identifying assumptions. This vignette walks through the
conceptual machinery one layer at a time and points to the canonical
references for each component.

```{r libs}
library(contagionchannels)
```

# 1. The detection-attribution distinction

A perennial confusion in the contagion literature is the elision of two
quite different statistical questions:

1.  **Detection.** *Does* country $i$'s return predict country $j$'s
    return beyond what is implied by their joint distribution at common
    factors?
2.  **Attribution.** *Through which channel* (trade, finance, sentiment,
    geopolitics, monetary policy) does that predictive content travel?

Detection is a local, model-free question best answered with a flexible
information-theoretic statistic; attribution is a structural question
that requires explicit identifying restrictions. Conflating the two
leads to the familiar problem in which a generic correlation spike is
read as evidence of a particular causal channel. The two-stage design
enforces a clean separation: the WQTE step asks only whether a flow
exists; the IV/LP/Rigobon step asks which structural shock the flow
embeds.

# 2. Stage 1: WQTE math intuition

The Stage 1 statistic combines three classical ingredients: Schreiber's
(2000) transfer entropy, the maximal-overlap discrete wavelet transform
(MODWT) for scale-localisation, and conditional quantile filtering for
tail sensitivity.

## Transfer entropy

For two stationary series $X_t$ and $Y_t$ Schreiber's transfer entropy
from $X$ to $Y$ is

$$
T_{X \to Y} \;=\; \sum p\!\left(y_{t+1}, y_t^{(k)}, x_t^{(\ell)}\right)
\; \log
\frac{p\!\left(y_{t+1} \mid y_t^{(k)}, x_t^{(\ell)}\right)}
     {p\!\left(y_{t+1} \mid y_t^{(k)}\right)},
$$

with $y_t^{(k)} = (y_t, y_{t-1}, \dots, y_{t-k+1})$ and similarly for
$x_t^{(\ell)}$. Intuitively, $T_{X \to Y}$ measures the bits per
observation that knowing the past of $X$ adds to the prediction of $Y$
above and beyond $Y$'s own past.

## Wavelet decomposition

We pre-filter both series with the MODWT (Daubechies LA8) at dyadic
scales $s \in \{1,2,3,4,5,6\}$, producing band-pass returns
$W_{X,t}^{(s)}$ that isolate fluctuations of period $[2^{s}, 2^{s+1}]$
trading days. Scale $s=5$ covers the 32-64 day band, which lines up with
the quarterly business-cycle horizon that motivates most of the channel
proxies.

## Quantile conditioning

Following Bekiros and co-authors, transfer entropy is computed
conditional on the source being in a quantile bin. Define
$X_t^{(\tau)} = \mathbf{1}\{X_t \le Q_X(\tau)\}$ and replace $X$ with
$X^{(\tau)}$ in the entropy expression above. We use $\tau = 0.50$ as
the default since the paper's primary interest is the typical (median)
flow, not just the tail.

## Bias correction

The WQTE point estimate $\widehat{T}^{(s,\tau)}_{i \to j}$ is
bias-corrected by Monte Carlo shuffling. We draw $B = 100$ permutations
of the source series, recompute the statistic, and subtract the mean.
Statistical significance is then assessed by comparing the corrected
statistic to the upper tail of the shuffled distribution.

```{r wqte-call, eval = FALSE}
F_mat <- compute_wqte_matrix(
  returns   = my_returns_xts,
  scale     = 5,
  tau       = 0.50,
  n_cores = 1,
)
```

References. Schreiber (2000) introduced transfer entropy
(<doi:10.1103/PhysRevLett.85.461>). The wavelet implementation borrows
from the `waveslim` package; quantile conditioning extends the Bekiros
et al. quantile-cross-spectral approach.

# 3. Stage 2: why multi-method identification?

A point estimate from any single estimator can be artefactual. Different
methods buy identification with different assumptions:

| Method | Identifying assumption | Failure mode |
|:---------------|:-------------------------------|:-----------------------|
| IV/2SLS | Channel-specific instruments are exogenous | Weak/invalid instruments |
| LASSO IV | Sparsity in the first stage | Approximate-sparsity violation |
| LP | Conditional mean linearity at each horizon | Non-linearity, regime change |
| Rigobon | Variance of structural shocks shifts across regimes | Insufficient variance shift |

The package reports all four in parallel. A **robust** finding is one in
which the dominant channel agrees across estimators; a **fragile**
finding is one that survives only under a single identifying assumption.

# 4. IV/2SLS with channel-specific instruments

Following the Stock-Watson (2018) external-instruments tradition, we use
one external proxy per channel:

$$
\widehat{F}_{i\to j,t} \;=\;
\alpha + \sum_{c=1}^{5} \beta_c \, C_{c,t} + u_{i\to j,t},
\qquad
C_{c,t} \;=\; \pi_c Z_{c,t} + v_{c,t}.
$$

The instruments $Z_{c,t}$ are: lagged Baltic Dry Index for **Trade**;
lagged FRA-OIS spread for **Financial**; lagged GPR-Daily for
**Geopolitical**; lagged VIX innovation for **Behavioral**; lagged
shadow-rate surprises for **Monetary_Policy**. Standard errors are
heteroskedasticity-robust and clustered at the directional-link level.
Over-identification is assessed via the Sargan-Hansen J-test; periods
with high rejection rates (GFC 67.3%, COVID 100%, ESDC 65.5%) are
reported but demoted to exploratory.

```{r iv-call, eval = FALSE}
fit <- iv_2sls_attribute(
  returns_period  = returns_pc,
  channels_period = channels_pc,
  links           = links_pc,
  instruments     = list(
    Trade           = "BDI_lag",
    Financial       = "FRAOIS_lag",
    Geopolitical    = "GPR_lag",
    Behavioral      = "VIX_innov_lag",
    Monetary_Policy = "ShadowRate_surp_lag"
  ),
  cluster_se = TRUE
)
```

Reference: Stock and Watson (2018) <doi:10.1111/ecoj.12593>.

# 5. LASSO instrument selection

When the candidate instrument list is long the
Belloni-Chernozhukov-Hansen (2014) post-LASSO IV estimator is preferred.
The first stage runs

$$
C_{c,t} = \pi_c' Z_t + v_{c,t},
$$

where $Z_t$ is a high-dimensional vector of candidate instruments
(macroeconomic surprises, policy-rate residuals, commodity shocks, etc.)
and $\pi_c$ is recovered via LASSO with the iteratively-tuned penalty
loadings of Belloni et al. The selected instruments feed the
second-stage 2SLS, and inference is conducted under the standard
sparsity conditions $s \log(p) / n \to 0$.

```{r lasso-call, eval = FALSE}
fit_lasso <- lasso_iv_attribute(
  returns_period  = returns_pc,
  channels_period = channels_pc,
  links           = links_pc,
  candidate_Z     = candidate_instrument_grid,
  selection       = "post_lasso"
)
```

Reference: Belloni, Chernozhukov and Hansen (2014)
<doi:10.1093/restud/rdt044>.

# 6. Local projections

Jordà (2005) projections estimate impulse responses by direct OLS at
each horizon $h$:

$$
\widehat{F}_{i\to j,t+h} \;=\;
\alpha_h + \beta_{c,h} \, C_{c,t} + \gamma_h' X_t + u_{i\to j,t+h},
\quad
h \in \{1, 5, 22\}.
$$

LP avoid the recursive bias of VAR-based IRFs and are robust to dynamic
mis-specification, at the cost of larger standard errors at long
horizons. We follow Stock-Watson and use Newey-West HAC standard errors
with bandwidth $h+1$.

```{r lp-call, eval = FALSE}
lp_fit <- local_projections(
  returns_period  = returns_pc,
  channels_period = channels_pc,
  links           = links_pc,
  horizons        = c(1, 5, 22),
  controls        = c("VIX_lag", "USD_lag")
)
```

Reference: Jordà (2005) <doi:10.1257/0002828053828518>.

# 7. Rigobon heteroskedasticity-based identification

When external instruments are unavailable but shock variances shift
across regimes, Rigobon (2003) provides identification through
heteroskedasticity. Partition the sample into a high-volatility regime
$H$ and a low-volatility regime $L$ (we use VIX terciles). The
reduced-form covariance matrices satisfy

$$
\Omega_H - \Omega_L = A \,(\Sigma_H - \Sigma_L)\, A',
$$

so that $A$ is identified up to sign from the difference in covariance
matrices, provided the structural shock-variance ratio differs across
regimes. The package's `rigobon_id()` function implements the GMM
estimator with the analytical Jacobian.

```{r rigobon-call, eval = FALSE}
rig_fit <- rigobon_id(
  returns_period  = returns_pc,
  channels_period = channels_pc,
  links           = links_pc,
  regime_split    = "vix_high_low"
)
```

Reference: Rigobon (2003) <doi:10.1162/003465303772815727>.

# 8. Cinelli-Hazlett robustness value

Identification-robust point estimates are not the end of the story:
unobserved confounders may still be lurking. The Cinelli-Hazlett (2020)
RV quantifies the minimum strength of an unobserved confounder,
expressed as its partial $R^2$ with both the treatment and the outcome,
that would push the estimated coefficient to zero (the "tipping point"
RV) or to its two-sided null (the "5% RV"):

$$
\mathrm{RV}_{q} \;=\;
\frac{1}{2}
\left(
\sqrt{f_q^{4} + 4 f_q^{2}} - f_q^{2}
\right),
\quad
f_q \;=\;
\frac{|\hat{\theta}| - q \cdot \widehat{\mathrm{SE}}}
     {|\hat{\theta}| + q \cdot \widehat{\mathrm{SE}}}\cdot
\sqrt{\mathrm{df}}.
$$

A coefficient with $\mathrm{RV} \ge 0.20$ is one that no plausible
single-confounder explanation can overturn under reasonable benchmarks.

```{r rv-call, eval = FALSE}
rv <- cinelli_hazlett_rv(
  theta = fit$shares,
  se    = fit$se,
  df    = fit$df_residual
)
```

Reference: Cinelli and Hazlett (2020) <doi:10.1111/rssb.12348>.

# 9. Identification-status classification

The decision rule used in the paper is intentionally conservative:

-   **Robust** — dominant channel agrees across IV/2SLS, LP-h5 and
    Rigobon, Sargan rejection rate below 50%, and Cinelli-Hazlett
    $\mathrm{RV} \ge
    0.20$ for the dominant channel.
-   **Fragile** — at least one of the three estimators disagrees on the
    dominant channel, *or* Sargan rejects in more than half of links,
    *or* $\mathrm{RV} < 0.20$.
-   **Exploratory** — Sargan rejection rate above 65% (GFC, COVID,
    ESDC). Reported but not used as primary evidence.

Out of the eight sub-periods, only **Pre-Crisis (Financial)** and **ESDC
(Financial)** clear the bar for *robust*; the remaining six are reported
as fragile, with the exploratory flag attached to GFC and COVID.

This explicit grading is what makes the package useful for empirical
work: the user is never invited to mistake a single-method point
estimate for a robust attribution claim. The codebase, vignettes, and
the report objects returned by `run_contagion_pipeline()` propagate the
identification status into every downstream summary table and figure.

# Session info

```{r session}
sessionInfo()
```
